# Setup

## Setting up R  

To get started, ensure you have a recent version or R and RStudio installed. 

### Step 1: Install R  

To install R head to [https://cran.rstudio.com/](https://cran.rstudio.com/) and follow the instructions for your operating system. 

### Step 2: Install RStudio  

Next, install RStudio Desktop IDE at [https://posit.co/download/rstudio-desktop/](https://posit.co/download/rstudio-desktop/).

### Packages  

To install the required packages, we run the `install.packages("packagename")` function.

```{r}
#| eval: false
# Install the required R packages for our analysis (first time use only)
install.packages("tidyverse")
install.packages("palmerpenguins")
install.packages("Metrics")
install.packages("car")
install.packages("skimr")
install.packages("AER")
```

:::{.callout-note}
The command `install.packages()` is only required the first time loading a new package or following any substantial updates. The `library()` command must be run  every time you start an R session. To save potential issues arising from unloaded packages, put any `library()` commands at the beginning of any script file.
:::

You should be able to now run the following commands: 

```{r}
#| eval: false
# Load the installed packages at the start of each session
library(tidyverse)
library(palmerpenguins)
library(Metrics)
library(car)
library(skimr)
library(AER)
```
